A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-10 paper.tex
Splice back in minimal eARG section

@@

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an
ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
% The final vital
% detail is that we associate a breakpoint with each recombination
% event.
% This information is sufficient to uniquely
% define the local genealogical trees at every position along the genome,
% which is the basic requirement for a data structure encoding
% genome-wide genetic ancestry.
As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
is usually concretely encoded by associating a recombination breakpoint
(which is null for common ancestor events) and an ordered list of parents
with each node. We refer to such a  data structure as an
``event ARG'' (eARG).
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure),
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
The order in which parents are listed at a recombination node is
therefore significant, telling us
from which parent the segments of genome to the left and right of the breakpoint
were inherited.

This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event older than a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic.

% What do we mean by an eARG, formally?
To be concrete, we can formally define the
classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(This is equivalent to having different ``types''
of node, where breakpoints are associated with recombination
nodes only).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
% For each recombination node $u$ we must
% have exactly two edges in $e$ such that $u$ is the child.
This encoding of an eARG is equivalent to the output
of programs such as
\texttt{ARGweaver}~\citep{rasmussen2014genome}
and \texttt{KwARG}~\citep{ignatieva2021kwarg},
and an example is given in Fig.~\ref{fig-arg-data-structure}A.
Note that the description above captures only the
graph topology: if we also wish to know the branch lengths we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

>>>>

\section*{Event ARGs}\label{eARG}
We are mainly interested in exploring the properties of gARGs in
this manuscript, but in order to contrast with the existing
methods we concretely define an ``event ARG'', or eARG.

In keeping with its original derivation in terms of a stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an
ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
% The final vital
% detail is that we associate a breakpoint with each recombination
% event.
% This information is sufficient to uniquely
% define the local genealogical trees at every position along the genome,
% which is the basic requirement for a data structure encoding
% genome-wide genetic ancestry.
As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
is usually concretely encoded by associating a recombination breakpoint
(which is null for common ancestor events) and an ordered list of parents
with each node. We refer to such a  data structure as an
``event ARG'' (eARG).
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure),
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
The order in which parents are listed at a recombination node is
therefore significant, telling us
from which parent the segments of genome to the left and right of the breakpoint
were inherited.

[TODO list the salient points that we want to contrast against
the gARG. We don't want this to seem like a straw man]

@@

% \section*{Event ARGs}\label{eARG}
% In keeping with its original derivation in terms of a stochastic
% process (see Appendix~\ref{sec-arg-history}), the classical definition of an
% ARG data structure is
% in terms of events.
% Nodes represent
% common ancestor or recombination events that occurred in the
% history of some set of sample genomes, and edges represent ancestral
% lineages~\citep{griffiths1996ancestral}.
% In a common ancestor event the inbound lineages are merged into a
% single ancestral lineage, and in a recombination
% event a lineage is split into two independent
% ancestral lineages.
% % The final vital
% % detail is that we associate a breakpoint with each recombination
% % event.
% % This information is sufficient to uniquely
% % define the local genealogical trees at every position along the genome,
% % which is the basic requirement for a data structure encoding
% % genome-wide genetic ancestry.
% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
% is usually concretely encoded by associating a recombination breakpoint
% (which is null for common ancestor events) and an ordered list of parents
% with each node. We refer to such a  data structure as an
% ``event ARG'' (eARG).
% To recover the local tree at genome coordinate $x$ (the most
% fundamental operation required for an ARG data structure),
% we traverse the graph rootwards from the leaves.
% At a particular
% node $u$, if it has one parent we are at a common ancestor
% node and we follow that parent. If we are at a
% recombination node, $u$ has two parents: if
% $x$ is less than the breakpoint we follow
% the edge to the first parent, and otherwise follow the edge
% to the second parent.
% The order in which parents are listed at a recombination node is
% therefore significant, telling us
% from which parent the segments of genome to the left and right of the breakpoint
% were inherited.
